Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-03-13 appendix.tex
generalized bcrit (attempt 1)
corrected generalized bcrit
updating

\section{Applicability of the vertically global shearing box}\label{global_corr}
In \S\ref{analytical} we used the vertically global shearing box
\citep[VGSB,][]{mcnally14} to make analytic progress.   
The main difference between the linear problem in the VGSB 
and the radially-local approximation
(Eq. \ref{lin_mass}---\ref{lin_energy}) adopted in our numerical study
is the absence of  terms associated with the radial disk structure in
the former framework. These terms appear in the energy, mass and radial momentum
equations and are reproduced here for convenience,

\begin{align}
  \ii \sigma W  &=c _s^2(r_0, z)\left( \left.\frac{\p\ln\rho}{\p
      z}\right|_0 \delta v_z + \frac{\gamma}{\Gamma} \frac{d\delta
    v_z}{dz}\right) + \left[\ii k_xc_s^2(r_0,z)
  \frac{\gamma}{\Gamma} + \hat{g}_c\left.\frac{1}{\rho}\frac{\p P}{\p
      r}\right|_0\right]\delta v_x  +
\frac{1}{t_c}\left(W-\frac{Q}{\Gamma}\right),\label{gcorr_terms1}\\
\ii \sigma Q &= c_s^2(r_0,z)\left(\ii k_x + \hat{g}_c
  \left.\frac{\p\ln{\rho}}{\p r}\right|_0\right)\delta v_x + c _s^2(r_0, z)\left( \left.\frac{\p\ln\rho}{\p
      z}\right|_0 \delta v_z + \frac{d\delta
    v_z}{dz}\right),\label{gcorr_terms2} \\
\ii \sigma \delta v_x & = \ii k_x W  -
\hat{g}_c\frac{1}{c_s^2(r_0,z)}\left.\frac{1}{\rho}\frac{\p P}{\p
  r}\right|_0Q - 2\Omega(r_0,z)\delta v_y,\label{gcorr_terms3}
\end{align}
where subscript $0$ denotes evaluation at $r=r_0$. We recall terms
proportional to $\hat{g}_c$ are associated with the background density
and pressure radial structures, and nominally $\hat{g}_c=1$.

@@

\subsection{Spurious growth of adiabatic perturbations in the VGSB}\label{analytic_adia} 
An important caveat of the VGSB is that it cannot be used in the
adiabatic limit. We explain this by considering vertically isothermal
disks and setting  $\beta\to\infty$ in 
Eq. \ref{vertiso_gov_nondim}, which can then be written as

\begin{align}
  0 =\dd v_z^{\prime\prime} + \left(1 + \ii \epsilon q
    \hat{k}\right)\left(\ln\rho^{\prime}\delta v_z\right)^\prime
  +\left\{\hat{\sigma}^2\left(\frac{1}{\gamma}+\hat{k}^2\right) 
    -\left(\frac{\gamma-1}{\gamma}\right)\left[\ln\rho^{\prime\prime}+\hat{k}^2\left(1-\frac{\ii\epsilon  
          q}{\hat{k}}\right)\ln\rho^{\prime 2}\right]\right\}\delta v_z.\label{adia_iso3}
\end{align}
We multiply Eq. \ref{adia_iso3} by $\rho\delta v_z^*$ and
integrate vertically, assume boundary terms vanish when integrating by
parts, to obtain
\begin{align}
  \hat{\sigma}^2\left(\frac{1}{\gamma} +
    \hat{k}^2\right)\int_{\zhat_1}^{\zhat_2}\rho|\delta
  v_z|^2 d\zhat 
  =&  \left(\frac{\gamma-1}{\gamma}\right)
  \int_{\zhat_1}^{\zhat_2}\rho|\delta v_z^\prime|^2 d\zhat
  +\frac{1}{\gamma}\int_{\zhat_1}^{\zhat_2}\frac{1}{\rho}|(\rho\delta
  v_z)^\prime|^2 d\zhat\notag\\
&+
  \left(\frac{\gamma-1}{\gamma}\right)\hat{k}^2\left(1-\frac{\ii\epsilon
      q}{\hat{k}}\right) \int_{\zhat_1}^{\zhat_2}\rho\ln\rho^{\prime
    2}|\delta v_z|^2 d\zhat
+ \ii\epsilon q \hat{k}
  \int_{\zhat_1}^{\zhat_2}\ln\rho^\prime(\rho\delta v_z^*)^\prime
  \delta v_z d\zhat.\label{adia_integral}
\end{align}
When $q\equiv0$, all the terms on the right-hand-side (RHS) are real. Then
$\hat{\sigma}^2$ is real, and  $\hat{\sigma}^2>0$ if $\gamma>1$. As
expected, a sub-adiabatically stratified disk is stable in the absence
of vertical shear.

However, in the presence of vertical shear $q\neq0$,
Eq. \ref{adia_integral} suggests $\hat{\sigma}$ is complex for real
$\khat$, which implies the possibility of instability, regardless of
how much $\gamma$ deviates from unity. This is contradicts the second 
Solberg-Hoiland criterion (Eq. \ref{solberg2}), which states that 
instability in a thin, adiabatic disk requires it to be close to 
neutrally stratified, i.e. $|\gamma-1|\ll1$ for vertically isothermal disks
evolved adiabatically. 
%also need real freq^2 in adia limit 
This is because the VGSB retains the vertical shear
($\p_z\Omega$) from the global disk, but ignores the global radial
disk structure responsible for it. Nevertheless, we demonstrate below
that this inconsistency is unimportant for the VSI because it occurs
for $\beta\ll1$.

\subsection{Effect of global radial gradients}
In  Fig. \ref{gcorr_compare}, we plot the effect of the global radial gradient terms
proportional to $\hat{g}_c$ in 
Eq. \ref{gcorr_terms1}---\ref{gcorr_terms3} by calculating the
fundamental VSI growth rates using three approaches. We compute growth rates from the dispersion
relation Eq. \ref{relax_disp}, which utilizes the VGSB;  from the
full linear problem (Eq. \ref{lin_mass}---\ref{lin_energy}) with
$\hat{g}_c=0$ (i.e. the linear VGSB equations); and from the full
linear problem with $\hat{g}_c=1$.

All three methods give similar behavior, and growth rates are in close
agreement for $\beta\lesssim 1$. Differences arise for 
$\beta\gtrsim1$, and as $\beta\to\infty$ the VGSB framework gives a
(spurious) growth rate as expected from the discussion
above. Inclusion of the global radial gradient terms results in the
correct behavior  ($\nu\to0$ as $\beta\to\infty$). Despite this
caveat, Fig. \ref{gcorr_compare} shows that provided we consider
$\beta\ll1$, then the VGSB framework is an adequate approximation.

@@

@@

\subsubsection{Stability without vertical shear}
In the absence of vertical shear $q=0$, Eq. \ref{simple_growth} can be
written as \begin{align}
  \hat{\sigma}^2\khat^2 =
  \left(L-\hat{\sigma}^2\right)\left(1-\hat{\sigma}^2\right). \label{iso_disp_full}
\end{align}
This is just the dispersion relation for axisymmetric waves in a
vertically isothermal disk already discussed in previous works 
\citep[e.g.][]{okazaki87,takeuchi98,tanaka02,zhang06,ogilvie13,barker14}. In
this case the solutions are Hermite polynomials, $W\propto
\He_L(\zhat)$.  The eigenfrequency $\hat{\sigma}$ is real so the disk
is stable, consistent with \S\ref{integral_relation}. The
low-frequency solution to Eq. \ref{iso_disp_full} correspond to
inertial waves \citep{balbus03}.

\subsubsection{Instability with vertical shear}
A simple solution to Eq. \ref{simple_growth} may be obtained by
considering small radial lengthscales $\khat\gg 1$. Then for the
low-frequency branch of Eq. \ref{simple_growth} 
\begin{align}
  \hat{\sigma}^2 \simeq L\left(\frac{1+\ii q \epsilon
       \hat{k}}{L+1+\hat{k}^2}\right).\label{simple_growth2}
\end{align}
The eigenfrequency $\hat{\sigma}$ is therefore complex for 
$L\geq1$, which signifies instability since we can choose
the sign of the square root to make $\hat{\nu} = \imag\hat{\sigma}>0$.

>>>>

\section{Applicability of the vertically global shearing box}\label{global_corr}
In \S\ref{analytical} we effectively employed the vertically global shearing box
\citep[VGSB,][]{mcnally14} to make analytic progress.   
The main difference between the linear problem in the VGSB 
and the radially-local approximation
(Eq. \ref{lin_mass}---\ref{lin_energy}) adopted in our numerical study
is the absence of terms explicitly associated with the radial disk structure in
the former framework. These terms appear in the energy, mass and radial momentum
equations and are labeled by $\hat{g}_c$ in
Eq. \ref{lin_mass}---\ref{lin_energy}. Nominally $\hat{g}_c=1$.

% \begin{align}
%   \ii \sigma W  &=c _s^2(r_0, z)\left( \left.\frac{\p\ln\rho}{\p
%       z}\right|_0 \delta v_z + \frac{\gamma}{\Gamma} \frac{d\delta
%     v_z}{dz}\right) + \left[\ii k_xc_s^2(r_0,z)
%   \frac{\gamma}{\Gamma} + \hat{g}_c\left.\frac{1}{\rho}\frac{\p P}{\p
%       r}\right|_0\right]\delta v_x  +
% \frac{1}{t_c}\left(W-\frac{Q}{\Gamma}\right),\label{gcorr_terms1}\\
% \ii \sigma Q &= c_s^2(r_0,z)\left(\ii k_x + \hat{g}_c
%   \left.\frac{\p\ln{\rho}}{\p r}\right|_0\right)\delta v_x + c _s^2(r_0, z)\left( \left.\frac{\p\ln\rho}{\p
%       z}\right|_0 \delta v_z + \frac{d\delta
%     v_z}{dz}\right),\label{gcorr_terms2} \\
% \ii \sigma \delta v_x & = \ii k_x W  -
% \hat{g}_c\frac{1}{c_s^2(r_0,z)}\left.\frac{1}{\rho}\frac{\p P}{\p
%   r}\right|_0Q - 2\Omega(r_0,z)\delta v_y,\label{gcorr_terms3}
% \end{align}
% where subscript $0$ denotes evaluation at $r=r_0$. We recall terms
% proportional to $\hat{g}_c$ are associated with the background density
% and pressure radial structures, and nominally $\hat{g}_c=1$.

@@

\subsection{Spurious growth of adiabatic perturbations in the VGSB}\label{analytic_adia} 
An important caveat of the VGSB is that it cannot be used in the
adiabatic limit. We explain this by considering vertically isothermal
disks and setting  $\beta\to\infty$ in 
Eq. \ref{vertiso_gov_nondim} (which itself is obtained by setting
$\hat{g}_c=0$ in Eq. \ref{lin_mass}---\ref{lin_energy}) to give 
\begin{align}
  0 =\dd v_z^{\prime\prime} + \left(1 + \ii \epsilon q
    \hat{k}\right)\left(\ln\rho^{\prime}\delta v_z\right)^\prime
  +\left\{\hat{\sigma}^2\left(\frac{1}{\gamma}+\hat{k}^2\right) 
    -\left(\frac{\gamma-1}{\gamma}\right)\left[\ln\rho^{\prime\prime}+\hat{k}^2\left(1-\frac{\ii\epsilon  
          q}{\hat{k}}\right)\ln\rho^{\prime 2}\right]\right\}\delta v_z.\label{adia_iso3}
\end{align}
We multiply Eq. \ref{adia_iso3} by $\rho\delta v_z^*$ and
integrate vertically, assume boundary terms vanish when integrating by
parts, to obtain
\begin{align}
  \hat{\sigma}^2\left(\frac{1}{\gamma} +
    \hat{k}^2\right)\int_{\zhat_1}^{\zhat_2}\rho|\delta
  v_z|^2 d\zhat 
  =&  \left(\frac{\gamma-1}{\gamma}\right)
  \int_{\zhat_1}^{\zhat_2}\rho|\delta v_z^\prime|^2 d\zhat
  +\frac{1}{\gamma}\int_{\zhat_1}^{\zhat_2}\frac{1}{\rho}|(\rho\delta
  v_z)^\prime|^2 d\zhat\notag\\
&+
  \left(\frac{\gamma-1}{\gamma}\right)\hat{k}^2\left(1-\frac{\ii\epsilon
      q}{\hat{k}}\right) \int_{\zhat_1}^{\zhat_2}\rho\ln\rho^{\prime
    2}|\delta v_z|^2 d\zhat
+ \ii\epsilon q \hat{k}
  \int_{\zhat_1}^{\zhat_2}\ln\rho^\prime(\rho\delta v_z^*)^\prime
  \delta v_z d\zhat.\label{adia_integral}
\end{align}
% When $q\equiv0$, all the terms on the right-hand-side (RHS) are real. Then
% $\hat{\sigma}^2$ is real, and  $\hat{\sigma}^2>0$ if $\gamma>1$. As
% expected, a sub-adiabatically stratified disk is stable in the absence
% of vertical shear.

In the presence of vertical shear $q\neq0$, Eq. \ref{adia_integral}
suggests $\hat{\sigma}$ is complex for real $\khat$, which implies the
possibility of instability, regardless of the value of
$|\gamma-1|$. This is contradicts the second Solberg-Hoiland criterion 
(Eq. \ref{solberg2}), which imply instability requires 
$|\gamma-1|\ll1$ for thin, vertically isothermal disks evolved adiabatically. 
%also need real freq^2 in adia limit 
This is because the VGSB retains the vertical shear
($\p_z\Omega$) from the global disk, but ignores the global radial
disk structure responsible for it. Nevertheless, we demonstrate below
that this inconsistency is unimportant for the VSI because it occurs
for $\beta\ll1$.

\subsection{Effect of global radial gradients}
In  Fig. \ref{gcorr_compare}, we plot the effect of the global radial gradient terms
proportional to $\hat{g}_c$ in Eq. \ref{lin_mass}---\ref{lin_energy}
by calculating the
fundamental VSI growth rates using three approaches. We compute growth rates from the dispersion
relation Eq. \ref{relax_disp}, which assumes $\hat{g}_c=0$;  from
Eq. \ref{lin_mass}---\ref{lin_energy} with $\hat{g}_c=0$; and from
Eq. \ref{lin_mass}---\ref{lin_energy}  with $\hat{g}_c=1$.

All three methods give similar behavior, and growth rates are in close
agreement for $\beta\lesssim 1$. Differences arise for 
$\beta\gtrsim1$, and as $\beta\to\infty$ the VGSB framework gives a
(spurious) growth rate as expected from the discussion
above. Inclusion of the global radial gradient terms results in the
correct behavior  ($\nu\to0$ as $\beta\to\infty$). Despite this
caveat, Fig. \ref{gcorr_compare} shows that provided we consider
$\beta\ll1$, then the VGSB framework is an adequate approximation to
study the VSI.

@@

\subsection{Limiting form with real frequencies and $\khat^2\gg 1$}\label{disp_neut_limit}
Assuming neutral modes with real $\hat{\sigma}=\hat{\omega}$ and $\khat^2\gg 1$, the real and imaginary parts of
Eq. \ref{relax_disp} are approximately 
\begin{align}
  0 =&M(M+1)(1-\epsilon^2 q^2 \khat^2) + 4\beta\epsilon q M(M+1)\left(\hat{\omega}\khat\right) 
 +\left[1 +
    \gamma\beta^2\left(1-\gamma\right)(1+2M)^2\right]\left(\hat{\omega}\khat\right)^2
 + 2\epsilon q \gamma\beta \left(\hat{\omega}\khat\right)^3 -  \left(\hat{\omega}\khat\right)^4\notag \\
  &+ \beta^2\gamma^2\hat{\omega}^2
  \left(\hat{\omega}\khat\right)^4,\label{relax_cond1} \\
   0=& 2\epsilon q M (M+1)\khat +
   \beta(1-\gamma)(1+2M)^2\khat\left(\hat{\omega}\khat\right) 
   + \epsilon q \khat \left(\hat{\omega}\khat\right)^2 -
   2\beta\hat{\omega}\left(\hat{\omega}\khat\right)^2
   - \epsilon q
   \gamma^2\beta^2\hat{\omega}\left(\hat{\omega}\khat\right)^3 
   +
   2\beta\gamma\hat{\omega}\left(\hat{\omega}\khat\right)^4 \label{relax_cond2}. 
\end{align}
These equations are used to derive a critical thermal timescale in
\S\ref{iso_vsi_beta_crit}. For fixed $M$ of $O(1)$,  we assume
$|\hat{\omega}\khat|$ remain constant but consider $\khat\gg 1$ so
that $|\hat{\omega}|\ll 1$.  Then for a thin disk $\epsilon \ll 1$
we neglect the second, fourth and last term in Eq. \ref{relax_cond1};
and neglect the last three terms in Eq. \ref{relax_cond2}. This
procedure gives Eq. \ref{relax_cond_simp1}---\ref{relax_cond_simp2},
but is not valid if $\beta$ or $|q|$ is large.

For the fundamental mode $M=0$ and Eq. \ref{relax_cond1} ---
\ref{relax_cond2} simply significantly. It is then straight
forward to eliminate $\hat{\omega}\khat$ between 
Eq. \ref{relax_cond_simp1} and \ref{relax_cond_simp2} to obtain
$\beta_\mathrm{crit}$ (Eq. \ref{iso_vsi_cond}) without the restrictive
assumption of $|\epsilon q \khat|\ll1$, as used in the main text to obtain the
equivalent expression for $M>0$ (Eq. \ref{bcrit_gen}).

%  In this case the fourth term
% in Eq. \ref{relax_cond1} may be retained but other aforementioned terms
% are still neglected. One then obtains an expression for $\beta$ by
% eliminating $\hat{\omega}\khat$ between Eq. \ref{relax_cond1} ---
% \ref{relax_cond2}. The expression for the critical thermal timescale
% for the fundamental mode, $\beta_\mathrm{crit}$
% (Eq. \ref{iso_vsi_con}, is then obtained by taking the thin disk
% limit.

%Taking the real and imaginary parts of Eq. \ref{relax_disp_fund} and considering 
% $\khat\gg1$, we find
% \begin{align}
%   0 =& \left[1 + \gamma\beta^2\left(1-\gamma\right)\right] + 2\epsilon q
%   \gamma\beta (\hat{\omega}\khat) -  (\hat{\omega}\khat)^2 \notag\\
%   &+ \beta^2\gamma^2\hat{\omega}^2 (\hat{\omega}\khat)^2,\label{relax_cond1}\\
%   0=& \beta(1-\gamma)\khat^2 + \epsilon q \khat^2 (\hat{\omega}\khat)
%   - 2\beta (\hat{\omega}\khat)^2 - \epsilon q \gamma^2\beta^2 (\hat{\omega}\khat)^3 \notag\\
%     &+ 2\beta\gamma(\hat{\omega}\khat)^4.  
% \label{relax_cond2}
% \end{align}
% This is a pair of simultaneous equations for
% $\hat{\omega}$ and $\beta$. Now, low-frequency axisymmetric waves in
% the absence of vertical shear are stable and  satisfy
% $\hat{\omega}\propto \khat^{-1}$ (Appendix \ref
% {stable_novshear}, Eq. \ref{iso_disp_full}).  We are thus motivated to
% seek solutions to Eq. \ref{relax_cond1}---\ref{relax_cond2} with  
% % we wish to seek a criterion independent of k 
% % \begin{align}
% %   \khat \to \infty, \quad \hat{\omega}\to 0 ,\quad \hat{\omega} \khat\text{ finite}.
% %\end{align}
% \begin{align}
%   \khat \gg1, \quad |\hat{\omega}|\ll 1 ,\quad
%   |\hat{\omega}\khat|=O(1). 
% \end{align}
% Then
% \begin{align}
%   \hat{\omega}\khat = \frac{(\gamma-1)\beta}{\epsilon q}
% \end{align}
% from Eq. \ref{relax_cond2}, which implies, from Eq. \ref{relax_cond1}
% that
% \begin{align}\label{beta_crit0}
%   \beta = \frac{1}{(\gamma-1)}\left[\frac{1}{\left(\epsilon
%         q\right)^2} - \frac{\gamma}{(\gamma-1)}\right]^{-1/2} 
% \end{align}

@@

\subsubsection{Stability without vertical shear}\label{stable_novshear}
In the absence of vertical shear $q=0$, Eq. \ref{simple_growth} can be
written as \begin{align}
  \hat{\sigma}^2\khat^2 =
  \left(L-\hat{\sigma}^2\right)\left(1-\hat{\sigma}^2\right). \label{iso_disp_full}
\end{align}
This is just the dispersion relation for axisymmetric waves in a
vertically isothermal disk already discussed in previous works 
\citep[e.g.][]{okazaki87,takeuchi98,tanaka02,zhang06,ogilvie13,barker14}. In
this case the solutions are Hermite polynomials, $W\propto
\He_L(\zhat)$.  The eigenfrequency $\hat{\sigma} = \hat{\omega}$ is real and the disk
is stable. The low-frequency branch of Eq. \ref{iso_disp_full} are 
inertial waves \citep{balbus03}. For 
$|\hat{\omega}|\ll 1$ and $L\geq 1$ the dispersion relation is
$\hat{\omega}^2\khat^2 = L$, or $\hat{\omega}\propto \khat^{-1}$ for
fixed $L$.  This inverse relation has been qualitatively
observed in numerical simulations of \cite{stoll14}.

\subsubsection{Instability with vertical shear}
A simple solution for $\hat{\sigma}$ may be obtained by balancing
the  last two terms of Eq. \ref{simple_growth} to obtain the
low-frequency branch,  
\begin{align}
  \hat{\sigma}^2 \simeq L\left(\frac{1+\ii q \epsilon
       \hat{k}}{L+1+\hat{k}^2}\right).\label{simple_growth2}
\end{align}
This signifies instability for $L\geq1$ 
since we can choose the sign of the square root to make $\hat{\nu} =
\imag\hat{\sigma}>0$.
